Exoplanet population inference and the rate of Earth analogs \\ from noisy, incomplete catalogs

### 2014-06-04 document/ms.tex
more discussion of extrapolation method

\newcommand{\Fig}[1]{Figure~\ref{fig:#1}}
\newcommand{\fig}[1]{\Fig{#1}}
\newcommand{\figlabel}[1]{\label{fig:#1}}
\newcommand{\Tab}[1]{Table~\ref{tab:#1}}
\newcommand{\tab}[1]{\Tab{#1}}
\newcommand{\tablabel}[1]{\label{tab:#1}}
\newcommand{\Eq}[1]{Equation~(\ref{eq:#1})}
\newcommand{\eq}[1]{\Eq{#1}}
\newcommand{\eqalt}[1]{Equation~\ref{eq:#1}}
\newcommand{\eqlabel}[1]{\label{eq:#1}}
\newcommand{\Sect}[1]{Section~\ref{sect:#1}}
\newcommand{\sect}[1]{\Sect{#1}}
\newcommand{\App}[1]{Appendix~\ref{sect:#1}}
\newcommand{\app}[1]{\App{#1}}
\newcommand{\sectlabel}[1]{\label{sect:#1}}

@@

Conditioned on our assumptions and the choices made in planet detection,
vetting and characterization pipeline (\citealt{petigura-a, petigura}), we
constrain the rate of small exoplanets orbiting Sun-like stars.
As part of this analysis we also place probabilistic constraints on the rate
of Earth analogs \gammaearth\ which we define as \emph{the number of planets
per star per natural logarithmic bin in period and radius, evaluated at the
period and radius of Earth} % Prof Tremaine, you owe us a beer.
\begin{eqnarray}
\gammaearth &=&
\left.\frac{\dd N}{\dd\ln\period\,\dd\ln\radius}\right|
_{\radius=\radius_\oplus,\,\period=\period_\oplus}\quad.
\end{eqnarray}
Since no Earth analogs have been detected, this constraint requires an
extrapolation in both period and radius.
\citet{petigura} performed this extrapolation by assuming that the period
distribution of planets in a small bin in radius is flat, obtaining
$\gammaearth \approx 0.12$.
We relax this assumption and extrapolate only by assuming that the occurrence
rate is a smooth function of period and radius; we find a much lower value for
\gammaearth.
We enforce the smoothness constraint by applying a flexible Gaussian process
regularization to the bin heights.

@@

\dfmtodo{Add more discussion here about how (because it is high variance) the
inv-det-eff method tends to find a lot of structure in the occurrence rate
function even when there shouldn't be.}

@@

\dfmtodo{Justify our choice of definition. Not habitability.}

As well as inferring the ocurrence distribution of exoplanets, this dataset
can also be used to constrain the rate of Earth analogs.
Explicitly, we constrain the rate of exoplanets \emph{evaluated at the
location of Earth}
\begin{eqnarray}
\gammaearth &=& \rate (\ln\period_\oplus,\,\ln\radius_\oplus) \\
&=&
\left.\frac{\dd N}{\dd\ln\period\,\dd\ln\radius}\right|
_{\radius=\radius_\oplus,\,\period=\period_\oplus}\quad.
\end{eqnarray}
That is, \gammaearth\ is the average number of exoplanets around a
``Sun-like'' star---per natural logarithmic bins in period and
radius---evaluated at the period and radius of Earth.
This is also the quantity that \citet{petigura} estimated but in different
units.

Since no Earth analogs have been found, any constraints on this rate must be
extrapolated from the existing observations.
This is generally done by assuming a functional form for the occurrence rate
function, constraining it using the observed candidates and extrapolating.
All published extrapolations are based on rigid models like a power law
(\citealt{catanzarite, traub}) or linear in the cumulative period distribution
(\citealt{petigura}).
All of these results are sensitive to this choice of function and to the other
implicit decisions (for example the size of bin that is considered Earth-like
by \citealt{petigura}).

We weaken the assumptions necessary for extrapolation by only assuming that
the distribution is smooth using the Gaussian process model described in
\sect{model}.
Under this model, the occurrence rate in bins where no objects have been
detected will be constrained by the height of nearby bins.

This model will, of course, break down if there is a sharp feature in the
occurrence rate function right at the location of Earth but otherwise this is
the most conservative extrapolation method published up to this point.

The results of our extrapolation method, when applied to the synthetic
catalogs described in the previous section, are shown in \fig{smooth-rate} and
\fig{simulation-rate}.
We implemented the extrapolation method from \citet{petigura}.

 that assumes a
linear relation in the cumulative period distribution for small planets and
compared these results---converting to the correct units---to our method and
the truth.
In both cases, our method returns a less precise but more accurate result for
the occurrence rate and the error bars given by the functional extrapolation
are overly optimistic.

@@

\Fig{real-results} shows posterior samples from the inferred occurrence rate
distribution as a function of period and radius conditioned on the catalog.
The marginalized distributions are qualitatively consistent with the
occurrence rate functions measured by \citet{petigura}.
The MCMC chain used to generate this figure is available online at
\dfmtodo{ADD SOME URL}.

>>>>

\newcommand{\figref}[1]{\ref{fig:#1}}
\newcommand{\Fig}[1]{Figure~\figref{#1}}
\newcommand{\fig}[1]{\Fig{#1}}
\newcommand{\figlabel}[1]{\label{fig:#1}}
\newcommand{\Tab}[1]{Table~\ref{tab:#1}}
\newcommand{\tab}[1]{\Tab{#1}}
\newcommand{\tablabel}[1]{\label{tab:#1}}
\newcommand{\Eq}[1]{Equation~(\ref{eq:#1})}
\newcommand{\eq}[1]{\Eq{#1}}
\newcommand{\eqalt}[1]{Equation~\ref{eq:#1}}
\newcommand{\eqlabel}[1]{\label{eq:#1}}
\newcommand{\Sect}[1]{Section~\ref{sect:#1}}
\newcommand{\sect}[1]{\Sect{#1}}
\newcommand{\App}[1]{Appendix~\ref{sect:#1}}
\newcommand{\app}[1]{\App{#1}}
\newcommand{\sectlabel}[1]{\label{sect:#1}}

@@

Conditioned on our assumptions and the choices made in planet detection,
vetting and characterization pipeline (\citealt{petigura-a, petigura}), we
constrain the rate of small exoplanets orbiting Sun-like stars.
As part of this analysis we also place probabilistic constraints on the rate
of Earth analogs \gammaearth\ which we define as \emph{the number of planets
per star per natural logarithmic bin in period and radius, evaluated at the
period and radius of Earth}
% A certain distinguished prof at the IAS owes us beer.
\begin{eqnarray}
\gammaearth &=&
\left.\frac{\dd N}{\dd\ln\period\,\dd\ln\radius}\right|
_{\radius=\radius_\oplus,\,\period=\period_\oplus}\quad.
\end{eqnarray}
Since no Earth analogs have been detected, this constraint requires an
extrapolation in both period and radius.
\citet{petigura} performed this extrapolation by assuming that the period
distribution of planets in a small bin in radius is flat, obtaining
$\gammaearth \approx 0.12$.
We relax this assumption and extrapolate only by assuming that the occurrence
rate is a smooth function of period and radius; we find a much lower value for
\gammaearth.
We enforce the smoothness constraint by applying a flexible Gaussian process
regularization to the bin heights.

@@

When applied to either simulated catalog, the inverse-detection-efficiency
method gives a high-variance estimate of the true occurrence rate.
One effect of this variance is that the inferred distribution will appear to
have more small-scale structure than the true underlying distribution.

@@

As well as inferring the occurrence distribution of exoplanets, this dataset
can also be used to constrain the rate of Earth analogs.
Explicitly, we constrain the occurrence rate of exoplanets orbiting
``Sun-like'' stars, as defined by \citet{petigura}, evaluated at the location
of Earth:
\begin{eqnarray}\eqlabel{gammaearth}
\gammaearth &=& \rate (\ln\period_\oplus,\,\ln\radius_\oplus) \\
&=&
\left.\frac{\dd N}{\dd\ln\period\,\dd\ln\radius}\right|
_{\radius=\radius_\oplus,\,\period=\period_\oplus}\quad.
\end{eqnarray}
That is, \gammaearth\ is the average number of exoplanets around a
Sun-like star---per natural logarithmic bins in period and radius---evaluated
at the period and radius of Earth.
This is also the quantity that \citet{petigura} estimated but in different
units.

In \eq{gammaearth}, we define ``Earth analog'' in terms of measurable
quantities with no mention of habitability or composition.
To some readers, this might seem an unsatisfying choice.
That being said, the composition of an exoplanet is notoriously difficult to
measure even with large uncertainty and any definition of habitability is
still extremely subjective.
With this in mind, we stick to the observable definition for this \paper.

Since no Earth analogs have been found, any constraints on this rate must be
extrapolated from the existing observations.
This is generally done by assuming a functional form for the occurrence rate
function, constraining it using the observed candidates and extrapolating.
All published extrapolations are based on rigid models of the occurrence rate
function (for example, a power law) fit to the catalog and evaluated at the
location of Earth (\citealt{catanzarite, traub}).
\citet{petigura} used their catalog of planet candidates to constrain the rate
of Earth analogs---using a definition similar to our definition in
\eq{gammaearth}---assuming an extremely rigid model of the occurrence rate
function: \emph{flat in logarithmic period}.
These results are sensitive to the choice of extrapolation function and the
specific definition of ``Earth analog''.

We weaken the assumptions necessary for extrapolation by only assuming that
the distribution is smooth using the Gaussian process regularization described
in \sect{model}.
Under this model, the occurrence rate in bins where no objects have been
detected will be constrained---with large uncertainty---by the heights of
nearby bins.
Therefore, even though there are no candidates that qualify as Earth analogs,
we simply fit our model of the occurrence rate function in a large enough
region of parameter space (including Earth) and compute the posterior
constraints on \gammaearth.
This works because the Gaussian process regularization actually captures our
prior beliefs about the shape of the occurrence rate function.
This model---and any other extrapolation---will, of course, break down if
there is an unmeasured sharp feature in the occurrence rate function near the
location of Earth and our method is the most conservative extrapolation
technique published to date.

For comparison, we also implemented and applied the extrapolation technique
applied by \citet{petigura}.
Their method assumes that, for small planets ($1 \le \radius/\radius_\oplus <
2$) on long periods ($\period > 50\,\mathrm{days}$), the occurrence rate flat
in logarithmic period or, equivalently, the cumulative rate is linear.
\citet{petigura} used the candidates in their catalog to estimate the slope of
the empirical cumulative period distribution and used that function to
extrapolate.
Instead of defining \gammaearth\ differentially, as we did in \eq{gammaearth},
\citet{petigura} constrained the integral of the occurrence rate over a box in
period and radius ($1 \le \radius/\radius_\oplus < 2$ and $200 \le
\period/\mathrm{day} < 400$).
Since their model implicitly assumes a constant occurrence rate across the
bin, the differential rate is just their number divided by the bin area.

Figures~\figref{smooth-rate} and~\figref{simulation-rate} compare our results
and the results of the \citet{petigura} extrapolation procedure when applied
to the synthetic catalogs.
Since these catalogs were simulated from a known population model, we know the
true value of \gammaearth\ and it is indicated in the Figures with a vertical
gray line.
In both cases, our method returns a less precise but more accurate result for
the occurrence rate and the error bars given by the functional extrapolation
are overly optimistic.
One major effect that leads to this bias is that the period distribution is
not flat.
Restricting the result to only include uniform models is equivalent to
applying an extremely informative prior that doesn't have enough freedom to
capture the complexity of the problem.
As a result, the posterior constraints on \gammaearth\ are dominated by this
prior choice and the resulting uncertainties are much smaller than they should
be.

@@

We directly applied our method to the \citet{petigura} sample and generated
MCMC samples from the posterior probability for the occurrence rate bin
heights, marginalizing over the hyperparameters of the Gaussian process model.
The resulting MCMC chain is available online at \dfmtodo{ADD SOME URL}.

\Fig{real-results} shows posterior samples from the inferred occurrence rate
distribution as a function of period and radius conditioned on the catalog.
The marginalized distributions are qualitatively consistent with the
occurrence rate functions measured using the inverse-detection-efficiency
method with larger uncertainties.

The period distribution integrated over various radius ranges is shown in
\fig{period}.
In agreement with \citet{dong}, we find that the period distribution of large
planets ($R > 8\,R_\oplus$) is inconsistent with the distribution of smaller
planets.
The occurrence rate of large planets appears to monotonically increase as a
function of log period while the distribution for small planets seems to turn
over at a relatively short period (around 50 days) and decrease for larger
periods.

The equivalent Figures for the radius distribution (integrated over period
bins) are shown in Figures~\figref{radius} and~\figref{linear-radius}.
